# Set up 

*For this workshop you will have available a VM*. However, if you which to use your local computer here are the set up instructions:

**IMPORTANT: If you have an M1 Mac - Make sure you have [gfortan](https://mac.r-project.org/tools/).**


## Recommended Computer Requirements:

System Requirements:

*Windows:*

* Windows 8.1 (64-bit) or later
* 4GB RAM
* SSD storage highly recommended
* Updated video/display drivers recommended


*macOS:*

* macOS 10.15 (Catalina) or later
* 4GB RAM
* SSD storage highly recommended

Install latest versions of:

* R
* RStudio


## Get the workshop material and data

In RStudio **create a new project**. This ensures all the files for this workshop are placed in their own folder.

Once you've created a new project, run the following R code

## Package Installation
 
For this workshop, several packages need to be installed.

BiocManager likes to update installed packages, but we have disabled this in the R code below. If your installation fails, then you might need to turn updates on. Note that if there are a large number of packages that BiocManager wants to update it can take several hours.

These instructions have been tested with R version 4.4.0 and Bioconductor version 3.16.

```{r eval=FALSE}
## Install required packages for Seurat and clustree:
install.packages(c("Seurat", "dplyr", "remotes", "R.utils", "harmony", 
                   "hdf5r", "clustree", "RColorBrewer","tidyverse","pander"))

## Install required Bioconductor packages
install.packages("BiocManager")
BiocManager::install(c('SingleR', 'celldex',
                       'BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment','edgeR' ),
                     update=FALSE)

# And the presto package from its github page using devtools
install.packages("devtools")
devtools::install_github('immunogenomics/presto@1.0.0')

```

## Raw Data

Raw data is provided on supplied VMs already.

```{r eval=FALSE}
# where are you? what folder are you working in
getwd()

# Navigate to the working directory
setwd("/your/R/project/directory")

## Download and untar the data
options(timeout=3600)
download.file(
    "https://bioinformatics.erc.monash.edu/home/lper0012/SingleCellWorkshopData/singlecell_2024/data.tar","data.tar")

untar("data.tar")

## Download workshop.R script. This file contains all the example code for this workshop.
download.file(
    "https://monashbioinformaticsplatform.github.io/scRNAseq_Workshop_ABACBS_2024/workshop.R","workshop.R")
```

