# Differential Expression {#de2}

[Slides](https://docs.google.com/presentation/d/1iTfkc2KgGy3d6yT9zCALZzk993pWwT9D/edit#slide=id.p1)

There are many different methods for calculating differential expression between groups in scRNAseq data. There are a number of review papers worth consulting on this topic.

There is also a good discussion of useing [pseudobulk approaches](http://bioconductor.org/books/3.15/OSCA.multisample/multi-sample-comparisons.html#creating-pseudo-bulk-samples) which is worth checking out if youre planning differential expression analyses.

We will be using a pseudobulk approach with limma. 

```{r }
head(seurat_object@meta.data)
```

How cells from each condition do we have?
```{r}
table(seurat_object$stim)
```

How many cells per individuals per group?
```{r}
table(seurat_object$ind, seurat_object$stim)
```
And for each sample, how many of each cell type has been classified?

```{r}
table(paste(seurat_object$ind,seurat_object$stim), seurat_object$cell)
```



## Prefiltering

#### Why do we need to do this? {- .rational}

If expression is below a certain level, it will be almost impossible to see any differential expression.

#### {-}


When doing differential expression, you generally ignore genes with low expression.
In single cell datasets, there are many genes like this. Filtering here to make our dataset smaller so it runs quicker, and there is less aggressive correction for multiple hypotheses.

How many genes before filtering?
```{r}
seurat_object
```


How many copies of each gene are there?
```{R}
total_per_gene <- rowSums(GetAssayData(seurat_object, assay='RNA', slot='counts'))
hist(log10(total_per_gene))
```



Lets keep only those genes with at least 50 copies across the entire experiment.
```{r}
seurat_object<- seurat_object[total_per_gene >= 50, ] 
```


How many genes after filtering?
```{r}
seurat_object
```



We might like to see the effect of IFN-beta stimulation on each cell type individually. For the purposes of this workshop, just going to test one cell type; CD14+ Monocytes 

An easy way is to subset the object. 

```{r}
# Set idents to 'cell' column.
Idents(seurat_object) <- seurat_object$cell
DimPlot(seurat_object, reduction = "umap_harmony")
seurat_object_celltype <- seurat_object[, seurat_object$cell == "CD14+ Monocytes" ]
DimPlot(seurat_object_celltype, reduction = "umap_harmony")
```




## Pseudobulk

Pseudobulk analysis is an option where you have biological replicates. It is essentially pooling the individual cell counts and treating your expreiment like a bulk RNAseq.


First, you need to build a pseudobulk matrix - the `AggregateExpression()` function can do this, once you set the 'Idents' of your seurat object to your grouping factor (here, thats a combination of individual+treatment called 'sample', instead of the 'stim' treatment column).

```{r}
# Tools for bulk differential expression
library(limma)
library(edgeR)


# Change idents to ind for grouping.
seurat_object_celltype$sample <- factor(paste(seurat_object_celltype$stim, seurat_object_celltype$ind, sep="_"))
Idents(seurat_object_celltype) <- seurat_object_celltype$sample

# THen pool together counts in those groups
# AggregateExperssion returns a list of matricies - one for each assay requested (even just requesting one)
pseudobulk_matrix_list <- AggregateExpression( seurat_object_celltype,  slot = 'counts', assays='RNA' )
pseudobulk_matrix      <- pseudobulk_matrix_list[['RNA']]
pseudobulk_matrix[1:5,1:4]
```

Now it looks like a bulk RNAseq experiment, so treat it like one.

We can use the popular `limma` package for differential expression. Here is one [tutorial](https://ucdavis-bioinformatics-training.github.io/2018-June-RNA-Seq-Workshop/thursday/DE.html), and the hefty reference manual is hosted by [bioconductor](https://bioconductor.org/packages/release/bioc/html/limma.html).

In brief, this code below constructs a linear model for this experiment that accounts for the variation in individuals and treatment. It then tests for differential expression between 'stim' and 'ctrl' groups.

```{R}
dge <- DGEList(pseudobulk_matrix)
dge <- calcNormFactors(dge)

# Remove _ or - and everything after it - yeilds stim group
stim <- gsub("[-_].*","",colnames(pseudobulk_matrix)) 

# Removing everything before the _ or - for the individual, then converting those numerical ind explictiy to text. Else limma will treat them as numbers!
ind  <- as.character(gsub(".*[-_]","",colnames(pseudobulk_matrix))) 

design <- model.matrix( ~0 + stim + ind)
vm  <- voom(dge, design = design, plot = FALSE)
fit <- lmFit(vm, design = design)

contrasts <- makeContrasts(stimstim - stimctrl, levels=coef(fit))
fit <- contrasts.fit(fit, contrasts)
fit <- eBayes(fit)

de_result_pseudobulk <- topTable(fit, n = Inf, adjust.method = "BH")
de_result_pseudobulk <- arrange(de_result_pseudobulk , adj.P.Val)
```


Look at the significantly differentially expressed genes:
```{r}
head(de_result_pseudobulk)
```


```{r}
p1 <- ggplot(de_result_pseudobulk, aes(x=AveExpr, y=logFC, col=adj.P.Val < 0.05)) +
  geom_point() +
  scale_colour_manual(values=c('TRUE'="red",'FALSE'="black")) + 
  theme_bw() +
  ggtitle("Pseudobulk")


p2 <- ggplot(de_result_pseudobulk, aes(x=logFC, y=-log10(P.Value), col=adj.P.Val < 0.05)) +
  geom_point() +
  scale_colour_manual(values=c('TRUE'="red",'FALSE'="black")) + 
  theme_bw() +
  ggtitle("Pseudobulk Test (Volcano)")

p1 + p2

```

#### Discussion {- .challenge}

These methods give different results. How would you decide which to use? How could you check an individual gene?

#### {-}



#### What about FindMarkers? {- .rational}

If you look at the seurat [differnetial expression vignette](https://satijalab.org/seurat/articles/de_vignette), you might have seen code that uses 'FindMarkers()' method to look for differential expression between groups.

By default, that fundtion uses a 'wilcox' test to look for differences in gene 
expression, treating _each individual cell as a replicate_.  In cases where biolocial replicates are simply not available this can be useful (e.g. pilot study, old public dataset or rare sample). 

However, when biological replicates are available, they should be used! 

#### {-}

