# (PART) Part 2 {-}

# Introduction to Part 2

## Part 1 Recap

In Part 1, we processed raw single-cell RNA-seq data through the following steps:

| Step | What we did | Why |
|---|---|---|
| Load data | Imported raw count matrices for 8 PBMC samples | Starting point — raw gene expression counts per cell |
| Quality control | Filtered low-quality cells based on count depth and mitochondrial gene percentage | Remove damaged or empty droplets that would skew downstream analysis |
| Normalisation | Normalised counts and identified highly variable genes | Make cells comparable despite differences in sequencing depth |
| Dimensionality reduction | Computed PCA and UMAP embeddings | Reduce noise and enable visualisation of structure in high-dimensional data |

## Clear your session

Before starting Part 2, clear your environment and restart your R session to remove any saved objects from Part 1.

In RStudio: **Session → Restart R** (or `Ctrl+Shift+F10` / `Cmd+Shift+F10` on Mac)

## Learning objectives

By the end of this lesson, you will be able to:

* Recall the experimental design and analysis goal
* Identify where this information is in the data

## Overview

Analysis goal: Identify molecular differences between stimulated vs. stimulated peripheral blood mononuclear cells (PBMCs).

PBMCs are stimulated in vitro with IFN-beta. This is expected to trigger the up or downregulation of immune pathways. In our data, we can measure this directly by comparing the gene expression levels of immune-related genes.

We have the following information in our Seurat object metadata:

-   `ind` identifies a cell as coming from one of 8 individuals.
-   `stim` identifies a cell as control or stimulated with IFN-beta.
-   `cell` contains the cell types identified by the creators of this data set.
-   `multiplets` classifies cells as singlet or doublet.

Read in libraries

```{r}
library(here)
library(qs2)
library(Seurat)
library(harmony)
library(tidyverse)
library(SingleCellExperiment)
library(SingleR)
library(celldex)
library(patchwork)
```

Read in the preprocessed Seurat object from part 1. We use `qs2` to efficiently load the data in. 

```{r}
seurat_object <- qs2::qs_read(here("data", "seurat_object_preprocessed.qs2"))
```

#### Exercise {- .challenge}

Use `names(seurat_object@meta.data)` to identify where this information is stored in the Seurat object.

```{r exercise1}
# Add your code here


```

#### {-}

#### Exercise {- .challenge}

Use the `table(seurat_object$<name>)` function to see the number of cells that belong to each metadata category. Replace `<name>` with the relevant metadata names (e.g. `ind`, `stim`)

```{r exercise2}
# Add your code here


```

#### {-}

It is useful to display the number of control and stimulated cells per individual

```{r}
table(seurat_object$ind, seurat_object$stim)
```

Takeaways:

-   Each cell in our Seurat object is annotated either as the control or stimualted cell
-   Each cell notes the individual the cells were derived from 
-   This is important so we know which cells to compare in the data
