# Introduction to Part 2

## Learning objectives

By the end of this lesson, you will be able to:

* Recall the experimental design and identify where this information is in the 
data
* Recall the analysis goal

## Overview

Analysis goal: Identify molecular differences between stimulated vs. stimulated peripheral blood mononuclear cells (PBMCs).

PBMCs are stimulated in vitro with IFN-beta. This is expected to trigger the up or downregulation of immune pathways. In our data, we can measure this directly by comparing the gene expression levels of immune-related genes.

We have the following information in our Seurat object metadata:

-   `ind` identifies a cell as coming from one of 8 individuals.
-   `stim` identifies a cell as control or stimulated with IFN-beta.
-   `cell` contains the cell types identified by the creators of this data set.
-   `multiplets` classifies cells as singlet or doublet.

-   Restart session - removes old saved data in the environment

Read in the preprocessed Seurat object from part 1:

```{r}
so <- readRDS("../data/seurat_object_preprocessed.rds")
```

EXERCISE: Use `names(so@meta.data)` to identify where this information is stored in the Seurat object.

```{r exercise1}
# Add your code here


```

EXERCISE: Use the `table(so$<name>)` function to see the number of cells that belong to each metadata category. Replace `<name>` with the relevant metadata names (e.g. `ind`, `stim`)

```{r exercise2}
# Add your code here


```

It is useful to display the number of control and stimulated cells per individual

```{r}
table(so$ctrl, so$stim)
```

Takeaways:

-   Each cell in our Seurat object is annotated either as the control or stimualted cell
-   Each cell notes the individual the cells were derived from 
-   This is important so we know which cells to compare in the data
