# Cluster Markers {#clustermarkers}

## Learning outcomes

- Explain how differentially expressed genes can be used as markers to characterise the biological identity of a cluster
- Run `FindAllMarkers` to identify positive markers for each cluster and visualise them
- Label clusters to known cell types using canonical marker genes

## Overview

**Why do we need to do this?**

We now have clusters, but we still do not know what they represent biologically. 
To interpret them, we need to identify which cell type each cluster represents.

Cells of the same type tend to co-express a characteristic set of genes. Genes that are highly and specifically expressed in one cluster compared to all others are its **markers**. By finding these differentially expressed genes and matching them against known cell type signatures, we can assign a biological identity to each cluster.

Note: in this example the cell types and their markers are well established in the literature, which makes this step straightforward. In practice, annotating novel or poorly characterised cell types requires careful expert curation.

## Finding differentially expressed features (cluster biomarkers)

The key parameters filter out genes that are unlikely to be informative: `min.pct` removes genes detected in very few cells, and `logfc.threshold` removes genes with small average differences between groups. This keeps the analysis focused on the most discriminatory markers.

Seurat can help you find markers that define clusters via differential expression. By default, it identifies positive and negative markers of a single cluster (specified in `ident.1`), compared to all other cells.  `FindAllMarkers()` automates this process for all clusters, but you can also test groups of clusters vs. each other, or against all cells.

```{r markers1, fig.height=8, fig.width=15}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(seurat_object, ident.1 = 2, min.pct = 0.25)
head(cluster2.markers, n = 5)
# find all markers distinguishing cluster 5 from clusters 0 and 3
cluster5.markers <- FindMarkers(seurat_object, ident.1 = 5, ident.2 = c(0, 3), min.pct = 0.25)
head(cluster5.markers, n = 5)
# find markers for every cluster compared to all remaining cells, report only the positive ones
seurat_object.markers <- FindAllMarkers(seurat_object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
seurat_object.markers %>% group_by(cluster) %>% slice_max(n = 2, order_by = avg_log2FC)
```

Seurat has several tests for differential expression which can be set with the test.use parameter (see our [DE vignette](de_vignette.html) for details). For example, the ROC test returns the 'classification power' `abs(AUC-0.5)*2` for any individual marker, ranging from 0 = random to 1 = perfect.

```{r markersroc, fig.height=8, fig.width=15}
cluster0.markers <- FindMarkers(seurat_object, ident.1 = 0, logfc.threshold = 0.25, test.use = "roc", only.pos = TRUE)
```

Two plots are most useful for confirming markers:

- `VlnPlot()` shows the distribution of expression across clusters. This is useful for checking that a marker is genuinely higher in one cluster and low elsewhere.
- `FeaturePlot()` overlays expression onto the UMAP. This isuseful for confirming that the distribution of the expression is restricted to the expected cluster.

```{r}
VlnPlot(seurat_object, features = c("MS4A1", "CD79A"))
# you can plot raw counts as well
VlnPlot(seurat_object, features = c("NKG7", "PF4"), layer = 'counts', log = TRUE)
```

```{r markerplots}
FeaturePlot(seurat_object, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A"))
FeaturePlot(seurat_object, features = c("FCGR3A", "LYZ", "PPBP", "CD8A"))
```

`DoHeatmap()` gives a cluster-level overview of the top markers across all clusters at once.

```{r clusterHeatmap, fig.height=12, fig.width=15}
maxcells=1500
top10 <- seurat_object.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
DoHeatmap(subset(seurat_object, downsample = maxcells), features = top10$gene) + NoLegend()
```

<details>
<summary>Additional visualisations</summary>

`RidgePlot()`, `CellScatter()`, and `DotPlot()` are alternative ways to inspect marker expression. They can be substituted for `FeaturePlot()` using the same feature arguments.

```{r ridgeplots}
RidgePlot(seurat_object, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A"))
RidgePlot(seurat_object, features = c("FCGR3A", "LYZ", "PPBP", "CD8A"))
```

For `CellScatter()`, you need the cell IDs of the two cells to compare. These can be found in the cell metadata.

```{r}
head(seurat_object@meta.data)

CellScatter(seurat_object, cell1 = "AGGGCGCTATTTCC-1", cell2 = "GGAGACGATTCGTT-1")
CellScatter(seurat_object, cell1 = "GGAGACGATTCGTT-1", cell2 = "TACGAGACCTATTC-1")
```

```{r}
DotPlot(seurat_object, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A", "FCGR3A", "LYZ", "PPBP", "CD8A"))
```

</details>

## Use markers to label or find a cluster

If you already know which markers define your cell types, you can score each cell for those markers directly using `AddModuleScore()`. This is useful for confirming the identity of a single cluster before committing to a full labelling. 

**Where to find canonical markers:**

- Literature searches for your tissue or cell population of interest
- [PanglaoDB](https://panglaodb.se/index.html) - curated cell type marker database
- [Human Protein Atlas](https://www.proteinatlas.org/) - gene expression across cell types and tissues

Fortunately in the case of this dataset, we can use canonical markers to easily match the unbiased clustering to known cell types:

| Markers       | Cell Type
|---------------|----------
| IL7R, CCR7    | Naive CD4+ T
| CD14, LYZ     | CD14+ Mono
| IL7R, S100A4  | Memory CD4+
| MS4A1         | B
| CD8A          | CD8+ T
| FCGR3A, MS4A7 | FCGR3A+ Mono
| GNLY, NKG7    | NK
| FCER1A, CST3  | DC
| PPBP          | Platelet

```{r}
genes_markers <- list(Naive_CD4_T = c("IL7R", "CCR7"))

seurat_object <- AddModuleScore(object = seurat_object, features = genes_markers, ctrl = 5, name = "Naive_CD4_T",
    search = TRUE)

# notice the name of the cluster has a 1 at the end
names(seurat_object@meta.data)

# label that cell type
seurat_object$cell_label = NA
seurat_object$cell_label[seurat_object$Naive_CD4_T1 > 1] = "Naive_CD4_T"
Idents(seurat_object) = seurat_object$cell_label

# plot
# Using a custom colour scale 
FeaturePlot(seurat_object, features = "Naive_CD4_T1", label = TRUE, repel = TRUE, reduction = "umap_harmony") + scale_colour_gradientn(colours = c("lightblue","beige","red"))
```

To score all cell types at once, define a named list and loop over it:

```{r}
cell_markers <- list(
    "Naive CD4+ T" = c("IL7R", "CCR7"),
    "CD14+ Mono"   = c("CD14", "LYZ"),
    "Memory CD4+"  = c("IL7R", "S100A4"),
    "B"            = c("MS4A1"),
    "CD8+ T"       = c("CD8A"),
    "FCGR3A+ Mono" = c("FCGR3A", "MS4A7"),
    "NK"           = c("GNLY", "NKG7"),
    "DC"           = c("FCER1A", "CST3"),
    "Platelet"     = c("PPBP")
)

for (cell_type in names(cell_markers)) {
    score_name <- gsub("[^A-Za-z0-9]+", "_", cell_type)
    seurat_object <- AddModuleScore(
        object   = seurat_object,
        features = list(cell_markers[[cell_type]]),
        name     = score_name,
        ctrl     = 5,
        search   = TRUE
    )
}

# Seurat appends 1 to each score name
grep("1$", names(seurat_object@meta.data), value = TRUE)
```

## Assigning cell type identity to clusters

Use the side-by-side plot below to match markers to cluster numbers: replace `"MS4A1"` in the `FeaturePlot` with a marker from the table, find the region of the UMAP that lights up, then read off the cluster number from the labelled DimPlot on the left.

```{r labelplot, fig.width=10}
DimPlot(seurat_object, group.by = "RNA_snn_res.0.5", reduction = "umap_harmony") |
  FeaturePlot(seurat_object, features = cell_markers$B, reduction = "umap_harmony")
```

Plot marker gene expression across all cell types:

```{r, fig.width=10}
library(patchwork)
for (cell_id in names(cell_markers)) {
  p <- FeaturePlot(seurat_object, features = cell_markers[[cell_id]], reduction = "umap_harmony")
  print(p + plot_annotation(title = cell_id))
}
```

Plot module scores across all cell types:

```{r, fig.width=10}
for (cell_type in names(cell_markers)) {
  score_name <- paste0(gsub("[^A-Za-z0-9]+", "_", cell_type), "1")
  p <- FeaturePlot(
      seurat_object,
      features  = score_name,
      label     = TRUE,
      repel     = TRUE,
      reduction = "umap_harmony"
    ) +
    scale_colour_gradientn(colours = c("lightblue", "beige", "red")) +
    labs(title = cell_type, colour = "Module\nscore")
  print(p)
}
```

## Exercise: Match cluster numbers with cell labels

For each cell type in the marker table, identify which cluster number it corresponds to and add it to the list below. Cluster 0 is done as an example.

```{r eval=FALSE}
Idents(seurat_object) <- seurat_object$RNA_snn_res.0.5

new.cluster.ids <- c(
    "0" = "Naive CD4+ T"
    # add the remaining clusters here
)

seurat_object <- RenameIdents(seurat_object, new.cluster.ids)
DimPlot(seurat_object, reduction = 'umap_harmony', label = TRUE, pt.size = 0.5) + NoLegend()
```

## save the plots

```{r save.img , eval=FALSE}

plot <- DimPlot(seurat_object, reduction = 'umap_harmony', label = TRUE, label.size = 4.5) + xlab("UMAP 1") + ylab("UMAP 2") + 
  theme(axis.title = element_text(size = 18), legend.text = element_text(size = 18)) + 
  guides(colour = guide_legend(override.aes = list(size = 10)))
ggsave(filename = "seurat_object3k_umap.jpg", height = 7, width = 12, plot = plot, quality = 50)
```

## save the seurat object

```{r save.rds, eval=FALSE}
saveRDS(seurat_object, file = "seurat_object3k_final.rds")
```

