# Cluster Markers {#clustermarkers}

## Learning outcomes

- Explain how differentially expressed genes can be used as markers to characterise the biological identity of a cluster
- Run `FindMarkers` to identify positive markers for each cluster and visualise them
- Label clusters to known cell types using canonical marker genes

## Overview

**Why do we need to do this?**

We now have clusters, but we still do not know what they represent biologically. 
To interpret them, we need to identify which cell type each cluster represents.

Cells of the same type tend to co-express a characteristic set of genes. Genes that are highly and specifically expressed in one cluster compared to all others are its **markers**. By finding these differentially expressed genes and matching them against known cell type signatures, we can assign a biological identity to each cluster.

Note: in this example the cell types and their markers are well established in the literature, which makes this step straightforward. In practice, annotating novel or poorly characterised cell types requires careful expert curation.

## Finding differentially expressed features (cluster biomarkers)

The key parameters filter out genes that are unlikely to be informative: `min.pct` removes genes detected in very few cells, and `logfc.threshold` removes genes with small average differences between groups. This keeps the analysis focused on the most discriminatory markers.

Seurat can help you find markers that define clusters via differential expression. By default, it identifies positive and negative markers of a single cluster (specified in `ident.1`), compared to all other cells.  `FindAllMarkers()` automates this process for all clusters, but you can also test groups of clusters vs. each other, or against all cells.

```{r}
# find all markers of cluster 0
cluster0.markers <- FindMarkers(seurat_object, ident.1 = 0, min.pct = 0.25)
head(cluster0.markers, n = 5)
```

```{r}
# find all markers distinguishing cluster 1 from local clusters 2, 5, 6
cluster1.markers <- FindMarkers(seurat_object, ident.1 = 1, ident.2 = c(2, 5, 6), min.pct = 0.25)
head(cluster1.markers, n = 5)
```

```{r}
# find markers for every cluster compared to all remaining cells, report only the positive ones
seurat_object.markers <- FindAllMarkers(seurat_object, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
seurat_object.markers %>% 
  group_by(cluster) %>% 
  slice_max(n = 2, order_by = avg_log2FC)
```

Two plots are most useful for confirming markers:

- `VlnPlot()` shows the distribution of expression across clusters. This is useful for checking that a marker is genuinely higher in one cluster and low elsewhere.
- `FeaturePlot()` overlays expression onto the UMAP. This is useful for confirming that the distribution of the expression is restricted to the expected cluster.

```{r}
seurat_object.markers %>% 
  group_by(cluster) %>% 
  slice_max(n = 4, order_by = avg_log2FC)
```

```{r}
cluster3_markers <- c("MS4A1", "CD79A", "IRF8", "CD83")
VlnPlot(seurat_object, features = cluster3_markers)
# you can plot raw counts as well
VlnPlot(seurat_object, features = cluster3_markers, layer = 'counts', log = TRUE)
```

```{r markerplots}
FeaturePlot(seurat_object, features = cluster3_markers, reduction = "umap_harmony")

cluster2_markers <- c("GZMA", "GNLY", "PRF1", "CLIC3")
FeaturePlot(seurat_object, features = cluster2_markers, reduction = "umap_harmony")
```

`DoHeatmap()` gives a cluster-level overview of the top markers across all clusters at once.

```{r clusterHeatmap, fig.height=12, fig.width=15}
maxcells=1500
top10 <- seurat_object.markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC)
DoHeatmap(subset(seurat_object, downsample = maxcells), features = top10$gene) + NoLegend()
```

## Top-down annotation with a marker table

`FindAllMarkers` gave us a ranked list of differentially expressed genes per cluster. To turn those rankings into cell-type labels, we need to match the genes against known markers from the literature. For a well-characterised tissue, the most efficient way to do this is to start from a curated list of canonical markers: genes whose expression patterns are already established for the cell types we expect.

We'll work through this as a three-step recipe.

### Step 1: Identify the expected cell types in your sample

Our dataset is peripheral blood mononuclear cells (PBMCs), a well-characterised mix of immune cells circulating in blood. In a PBMC sample you should expect to find:

- CD4+ T cells (helper T cells)
- CD8+ T cells (cytotoxic T cells)
- B cells
- NK (natural killer) cells
- CD14+ monocytes (classical)
- FCGR3A+ monocytes (non-classical)
- Dendritic cells
- Megakaryocytes / platelets

Knowing this up front matters because it tells you which markers to look up in Step 2, and what to expect when you visualise the data in Step 3.

### Step 2: Look up canonical markers for each cell type

Canonical markers are genes whose specific expression in a given cell type is well-established. Where to find them:

- [PanglaoDB](https://panglaodb.se/index.html): a curated cell-type marker database
- [Human Protein Atlas](https://www.proteinatlas.org/): gene expression across cell types and tissues
- Published literature for your tissue or cell population

**Worked example: confirming NK markers using the Human Protein Atlas**

Cluster 2's top markers from `FindAllMarkers` above included `GNLY`, `GZMA`, and `PRF1`. Let's verify these point to NK cells.

1. Open [https://www.proteinatlas.org/](https://www.proteinatlas.org/)
2. Search for `GNLY`
3. Click the **Single cell** tab
4. Look at the cell type group specificity plot. `GNLY` should show highest expression in NK cells.
5. Cross-check with one or two more candidates (`NKG7`, `KLRD1`).
6. Record the markers you'll use for NK cells in R:

```{r}
nk_markers <- c("NKG7", "GNLY", "KLRD1")
```

In your own work, you would repeat this lookup for every expected cell type. For time, we've collated the result for all eight PBMC cell types below.

### Step 3: Build the marker list in R and visualise

This is the named list you'd build from Step 2, mapping each expected cell type to its canonical markers:

```{r}
cell_markers <- list(
  "CD4 T cells"       = c("CD4", "IL7R"),
  "CD8 T cells"       = c("CD8A", "CD8B"),
  "B cells"           = c("MS4A1", "CD79A"),
  "NK cells"          = c("NKG7", "GNLY", "KLRD1"),
  "CD14+ Monocytes"   = c("CD14", "LYZ", "S100A9"),
  "FCGR3A+ Monocytes" = c("FCGR3A", "MS4A7"),
  "Dendritic cells"   = c("CST3", "FCER1A"),
  "Megakaryocytes"    = c("PF4", "PPBP")
)
```

The most informative plot for annotation is a `DotPlot`: a compact cluster by gene grid showing both the proportion of expressing cells (dot size) and average expression (colour). It lets us read off all markers and all clusters at once.

```{r, fig.width=12}
all_markers <- unique(unlist(cell_markers))

# Filter to genes actually present in the assay (some symbols may not be in this dataset)
features_to_plot <- all_markers[all_markers %in% rownames(seurat_object)]

DotPlot(seurat_object, features = features_to_plot) + RotatedAxis()
```

Read the plot column by column: for each marker group, find the cluster (row) where the dots are largest and darkest. That cluster expresses those markers and is likely that cell type.

## Exercise: Match cluster numbers with cell labels

Use the `DotPlot` above to assign a cell-type label to each cluster. Cluster 0 is done as an example.

**Workflow for each blank cluster:**

1. Look at the `DotPlot`. Find the row (cluster) where one marker group's dots are largest and darkest.
2. Cross-check with a `FeaturePlot` for that group's markers (code provided below). Does the cluster on the UMAP light up?
3. If you're unsure between two candidate cell types, use `AddModuleScore` (next section) to score every cell against each candidate signature.
4. Fill in your answer in `new.cluster.ids`.

DISCUSS: One cluster will be harder to call than the others. Which is it, and what made it ambiguous?

```{r}
Idents(seurat_object) <- seurat_object$RNA_snn_res.0.6

new.cluster.ids <- c(
  "0" = "CD4 T cells",   # done as an example
  "1" = "",
  "2" = "",
  "3" = "",
  "4" = "",
  "5" = "",
  "6" = "",
  "7" = ""
)
```

Optional: inspect the spatial distribution of any cell type's markers on the UMAP. Change `"NK cells"` to the cell type you want to check:

```{r, eval=FALSE}
FeaturePlot(seurat_object, features = cell_markers[["NK cells"]], reduction = "umap_harmony")
```

Apply your labels and visualise the annotated UMAP:

```{r}
seurat_object <- RenameIdents(seurat_object, new.cluster.ids)
DimPlot(seurat_object, reduction = "umap_harmony", label = TRUE, pt.size = 0.5) + NoLegend()
```

## Verifying a cluster's identity with `AddModuleScore`

Visual inspection of a `DotPlot` is fast, but for ambiguous clusters it helps to get a single quantitative score per cell. `AddModuleScore` computes the average expression of a marker gene set, minus the average expression of a random set of control genes with similar overall expression levels. The result is a relative score: positive when a cell expresses the marker set more than expected by chance, and robust to overall expression magnitude.

We'll use it here to verify that cluster 2 really is an NK-cell cluster.

```{r}
# Reset Idents back to cluster numbers so we can refer to them by number
Idents(seurat_object) <- seurat_object$RNA_snn_res.0.6

seurat_object <- AddModuleScore(
  object   = seurat_object,
  features = list(cell_markers[["NK cells"]]),
  ctrl     = 5,
  name     = "NK_cell"
)

# AddModuleScore appends a "1" to the name (NK_cell1)
names(seurat_object@meta.data)
```

Plot the score on the UMAP. Cells in cluster 2 should have the highest NK signature scores:

```{r}
FeaturePlot(
  seurat_object, features = "NK_cell1",
  label = TRUE, repel = TRUE, reduction = "umap_harmony"
) + scale_colour_gradientn(colours = c("lightblue", "beige", "red"))
```

Compare against the unsupervised clusters:

```{r}
DimPlot(seurat_object, reduction = "umap_harmony", group.by = "RNA_snn_res.0.6", label.box = TRUE, label = TRUE)
```

DISCUSS: Which clusters are you least confident in? What would you do to resolve them: repeat the module score for a different candidate cell type, or try a different approach entirely? In the next chapter we'll see one alternative, reference-based annotation with SingleR.