# SingleR

## Learning outcomes

- Apply bioconductor packages to automate cell type annotations
- Evaluate the trade-offs between reference-based and marker-based annotation

## Overview

**Why do we need to do this?**

In the previous section we annotated clusters manually using canonical marker genes. This works well for well-characterised cell types, but it is slow, subjective, and depends on the analyst's domain knowledge. Different researchers may reach different conclusions from the same data.

Reference-based annotation automates this process by comparing each cell's expression profile against a curated reference dataset of known cell types. This is faster, reproducible, and less dependent on prior knowledge, making it easier to annotate large datasets or cell types you are less familiar with.

Read in our manually annotated seurat object.

```{r}
seurat_object <- qs2::qs_read("../data/seurat_object_preprocessed.harmony.clustered.markers.qs2")
Idents(seurat_object) <- seurat_object$markers
```

In this workshop we have focused on the Seurat package.  However, there is another whole ecosystem of R packages for single cell analysis within Bioconductor.  We won't go into any detail on these packages in this workshop, but there is good material describing the object type online : [OSCA](https://robertamezquita.github.io/orchestratingSingleCellAnalysis/data-infrastructure.html).

For now, we'll just convert our Seurat object into an object called SingleCellExperiment. This allows us to utilise the Bioconductor packages.

```{r}
sce <- as.SingleCellExperiment(seurat_object)
sce
```

We will now use a package called SingleR to label each cell.  SingleR uses a reference data set of cell types with expression data to infer the best label for each cell. A convenient collection of cell type reference is in the `celldex` package which currently contains the follow sets:

```{r}
ls('package:celldex')
```

We will use `HumanPrimaryCellAtlasData` because our dataset is human PBMCs - broad, well-characterised immune cell types that are well represented in this atlas. It provides both high-level (`label.main`) and fine-grained (`label.fine`) annotations.

This reference would be a poor choice if your data came from a specific tissue (e.g. brain, liver), a non-human species, or contained rare or novel cell types not present in the atlas (e.g. tumour cells). In those cases, you would need a more appropriate reference. For example a tissue-specific atlas or one built from your own annotated data.

Let's download the reference dataset:

```{r}
# This too is a sce object,
# colData is equivalent to seurat's metadata
ref.set <- celldex::HumanPrimaryCellAtlasData()
```

HPCA contains two categories of labels at different granularities, `main` and `fine`. Display the `main` labels.

```{r}
unique(ref.set$label.main)
```

An example of the types of "fine" labels.

```{r}
head(unique(ref.set$label.fine), n = 20)
```

Now we'll label our cells using the SingleCellExperiment object, with the above reference set. This will take a minute.

```{r}
pred.cnts <- SingleR::SingleR(test = sce, ref = ref.set, labels = ref.set$label.main)
```

Labels assigned to fewer than 10 cell are likely low-confidence predictions - too few cells to represent a genuine population. We collapse these into an "Other" category and transfer the remaining labels back to the Seurat object.

```{r}
lbls.keep <- table(pred.cnts$labels)>10
seurat_object$SingleR.labels <- ifelse(lbls.keep[pred.cnts$labels], pred.cnts$labels, 'Other')
DimPlot(seurat_object, reduction='umap_harmony', group.by='SingleR.labels')
```

### Comparing annotation methods

Before comparing, it is worth summarising what each approach offers:

|             | Reference-based (SingleR)               | Marker-based (manual)     |
|-------------|-----------------------------------------|---------------------------|
| Speed       | Fast, automated                         | Slow, manual              |
| Objectivity | Reproducible                            | Analyst-dependent         |
| Flexibility | Bounded by reference                    | Uses any domain knowledge |
| Limitations | Misses cell types absent from reference | Subjective, error-prone   |

The `$cell` column in our metadata contains the cell type labels provided by the original dataset authors. We can use these as a ground truth to assess both annotation approaches side by side.

```{r, fig.width=16, fig.height=5}
library(patchwork)
p1 <- DimPlot(seurat_object, reduction = "umap_harmony", label = TRUE, repel = TRUE) +
        ggtitle("Marker-based") + NoLegend()
p2 <- DimPlot(seurat_object, reduction = "umap_harmony", group.by = "SingleR.labels",
              label = TRUE, repel = TRUE) +
        ggtitle("SingleR (HPCA main)") + NoLegend()
p3 <- DimPlot(seurat_object, reduction = "umap_harmony", group.by = "cell",
              label = TRUE, repel = TRUE) +
        ggtitle("Ground truth") + NoLegend()
p1 | p2 | p3
```

To go beyond the visual and count exact agreement, compare SingleR labels against the ground truth:

```{r}
table(SingleR = seurat_object$SingleR.labels, Ground_truth = seurat_object$cell)
```

ANSWER IN SLACK: Looking at the three plots and the table above, where do manual annotation, SingleR, and the ground truth agree? Where do they differ? What might explain the discrepancies?

## Exercise: Fine-grained annotation

Rerun SingleR using `label.fine` instead of `label.main` and plot the results.

```{r eval=FALSE}
# Add your code here


```

DISCUSS: How does the granularity of the labels change? Does finer annotation make the results easier or harder to interpret? What trade-off does this introduce?

Fine-grained labels capture biologically meaningful distinctions that broad labels miss — for example, separating CD4+ naive from CD4+ memory T cells. However, the more granular the labels, the harder it is for the algorithm to distinguish them reliably: closely related subtypes have very similar expression profiles, and the reference may not have enough examples of each subtype to anchor them confidently. In practice, run both: use `label.main` to confirm the broad populations are correct, then apply `label.fine` to any cluster where a finer distinction matters for your biological question.

```{r, eval=F}
qs2::qs_save(seurat_object, "../data/seurat_object_preprocessed.harmony.clustered.markers.singler.qs2")
```

