# Differential Expression

## Learning outcomes

- Explain why pseudobulk approaches treat each cell identity and condition as the unit of analysis, rather than each cell 
- Interpret differential expression (DE) results to identify genes that change in response to a condition
- Assess whether DE results are cell-type specific or broadly shared across the dataset 

## Overview

**Why do we need to do this?**

[Slides](https://docs.google.com/presentation/d/1iTfkc2KgGy3d6yT9zCALZzk993pWwT9D/edit#slide=id.p1)

There are many different methods for calculating differential expression between groups in scRNAseq data. There are a number of review papers worth consulting on this topic.


We will be using a pseudobulk approach with limma. There is also a good discussion of using [pseudobulk approaches here](http://bioconductor.org/books/3.15/OSCA.multisample/multi-sample-comparisons.html#creating-pseudo-bulk-samples), which is worth checking out if you're planning differential expression analyses.


## Experimental question and annnotation

In this (NB: modified!) experimental data taken from [Kang et al 2018 - Multiplexed droplet single-cell RNA-sequencing using natural genetic variation](https://pubmed.ncbi.nlm.nih.gov/29227470/), there are IFN-beta stimulated and unstimulated control PBMCs from 8 different lupus patient samples.

For this example, we'd like to look at the gene expression changes that occur with IFN-beta stimulation in the CD14+ Monocytes.

To recap the annotation in our object

*  **ind** identifies a cell as coming from one of 8 individuals.
* **stim** identifies a cell as control or stimulated with IFN-beta.
* **cell** contains the cell types identified by the creators of this data set.



So - How many cells from each condition do we have?

```{r}
table(seurat_object$stim)
```

How many cells per individuals per group?
```{r}
table(seurat_object$ind, seurat_object$stim)
```
And for each sample, how many of each cell type has been classified?

```{r}
table(paste(seurat_object$ind,seurat_object$stim), seurat_object$cell)
```


Lets add some columns to our metadata - with a unique value per sample and per sample+celltype.

```{r}
seurat_object$sample_name          <- paste(seurat_object$stim, seurat_object$ind, sep=".")
seurat_object$sample_celltype_name <- paste(gsub(" ","",seurat_object$cell), seurat_object$ind, seurat_object$stim, sep=".")
```

## Prefiltering

#### Why do we need to do this? {- .rational}

If expression is below a certain level, it will be almost impossible to see any differential expression.

#### {-}


When doing differential expression, you generally ignore genes with low expression.
In single cell datasets, there are many genes like this. Filtering here to make our dataset smaller so it runs quicker, and there is less aggressive correction for multiple hypotheses.

How many genes before filtering?
```{r}
seurat_object
```


How many copies of each gene are there?
```{R}
total_per_gene <- rowSums(GetAssayData(seurat_object, assay='RNA', layer='counts'))
hist(log10(total_per_gene))
```



Lets keep only those genes with at least 50 copies across the entire experiment.
```{r}
seurat_object<- seurat_object[total_per_gene >= 50, ] 
```


How many genes after filtering?
```{r}
seurat_object
```



We might like to see the effect of IFN-beta stimulation on each cell type individually. For the purposes of this workshop, just going to test one cell type; CD14+ Monocytes 

An easy way is to subset the object. 

```{r}
# Set idents to 'cell' column.
Idents(seurat_object) <- seurat_object$cell
DimPlot(seurat_object, reduction = "umap_harmony")
seurat_object_celltype <- seurat_object[, seurat_object$cell == "CD14+ Monocytes" ]
DimPlot(seurat_object_celltype, reduction = "umap_harmony")
```




## Pseudobulk

Pseudobulk analysis is an option where you have biological replicates. It is essentially pooling the individual cell counts and treating your experiment like a bulk RNAseq.


First, you need to build a pseudobulk matrix - the `AggregateExpression()` function can do this, once you set the 'Idents' of your seurat object to your grouping factor (here, that's that combination of individual+treatment called 'sample', instead of the 'stim' treatment column).

```{r}
# Tools for bulk differential expression
library(limma)
library(edgeR)


# Pull out a table of our sample information - the sample, individual, stim condition and celltype
pseudobulk_annotation_table <- FetchData(seurat_object_celltype, vars = c('sample_name', 'ind', 'stim', 'cell', 'sample_celltype_name')) |> 
  as_tibble()  |>
  group_by(across(everything()))  |> # group by unique row combinations
  summarise(n_cells=n())             # and count the number of cells

# Change idents to our samples for grouping.
Idents(seurat_object_celltype) <- seurat_object_celltype$sample_name

# Then pool together counts in those groups
# AggregateExperssion returns a list of matrices - one for each assay requested (even just requesting one)
pseudobulk_matrix_list <- AggregateExpression( seurat_object_celltype,  slot = 'counts', assays='RNA' )
pseudobulk_matrix      <- pseudobulk_matrix_list[['RNA']]
pseudobulk_matrix[1:5,1:4]


# Match the order of the matrix with the annotation. It did already match in this case, but a mismatch would cause silent swapping of the data!
pseudobulk_matrix <- pseudobulk_matrix[,pseudobulk_annotation_table$sample_name]
# And double check our annotation matches our data
# Ie. check that all pseudobulk matrix column labels match in the same order of the 'sample_name' column of the annotation table. 
stopifnot( all(colnames(pseudobulk_matrix) == pseudobulk_annotation_table$sample_name ))  


```

Now it looks like a bulk RNAseq experiment, so treat it like one.

## Test differential expression

We can use the popular `limma` package for differential expression. Here is one [tutorial](https://ucdavis-bioinformatics-training.github.io/2018-June-RNA-Seq-Workshop/thursday/DE.html), and the hefty reference manual is hosted by [bioconductor](https://bioconductor.org/packages/release/bioc/html/limma.html).

In brief, this code below constructs a linear model for this experiment that accounts for the variation in individuals and treatment. It then tests for differential expression between 'stim' and 'ctrl' groups.

```{R}
dge <- DGEList(pseudobulk_matrix)
dge <- calcNormFactors(dge)

# Extract the individuals (as text, do not want them treated numerically!) and stim condition from our pseudobulk annotation table
ind  <- as.character(pseudobulk_annotation_table$ind) 
stim <- pseudobulk_annotation_table$stim

design <- model.matrix( ~0 + stim + ind)
vm  <- voom(dge, design = design, plot = FALSE)
fit <- lmFit(vm, design = design)

contrasts <- makeContrasts(stimstim - stimctrl, levels=coef(fit))
fit <- contrasts.fit(fit, contrasts)
fit <- eBayes(fit)

de_result <- topTable(fit, n = Inf, adjust.method = "BH")
de_result <- arrange(de_result , adj.P.Val)
```


Look at the significantly differentially expressed genes:
```{r}
head(de_result)
```


```{r}
p1 <- ggplot(de_result, aes(x=AveExpr, y=logFC, col=adj.P.Val < 0.01)) +
  geom_point(pch=3, alpha=0.5) +
  scale_colour_manual(values=c('TRUE'="red",'FALSE'="black")) + 
  theme_bw() +
  ggtitle("Pseudobulk")


p2 <- ggplot(de_result, aes(x=logFC, y=-log10(P.Value), col=adj.P.Val < 0.01)) +
  geom_point(pch=3, alpha=0.5) +
  scale_colour_manual(values=c('TRUE'="red",'FALSE'="black")) + 
  theme_bw() +
  ggtitle("Pseudobulk Test (Volcano)")

p1 + p2

```

## Plot gene expression

Its always a good idea to look at the data. Lets plot the expression of a differentially expressed gene, across all celltypes.


We will build another expression matrix for this for two reasons: 

1) Need normalised counts: We used counts for the differential expression. But counts are unnormalised, so we need a normalised version for visually comparing samples.
2) Plot all celltypes: We used just monocytes for the comparison, but we'd like to plot all the celltypes. 


This code will build another counts matrix with a 'sample' for every combination of sample + celltype. Plus a table of annotation to go with it. 



```{r}
# Pull out a table of our sample information - the sample, individual, stim condition and celltype, for all celltypes
pseudobulk_annotation_table.allcelltypes <- FetchData(seurat_object, vars = c('sample_name', 'ind', 'stim', 'cell', 'sample_celltype_name'))  %>% as_tibble() %>%
  group_by(across()) %>% 
  summarise(n_cells=n())

pseudobulk_annotation_table.allcelltypes
```

Remove anything with less than 20 cells (we didn't need to bother for the CD14+ Monocytes)

```{r}
pseudobulk_annotation_table.allcelltypes <- filter(pseudobulk_annotation_table.allcelltypes, n_cells >= 20)
```


```{r}
## And all celltype counts matrix
Idents(seurat_object) <- seurat_object$sample_celltype_name

pseudobulk_matrix_list.allcelltypes <- AggregateExpression( seurat_object,  slot = 'counts', assays='RNA' )
pseudobulk_matrix.allcelltypes      <- pseudobulk_matrix_list.allcelltypes[['RNA']]

# Match the order of the matrix with the annotation. It did already match in this case, but a mismatch would cause silent swapping of the data!
pseudobulk_matrix.allcelltypes <- pseudobulk_matrix.allcelltypes[,pseudobulk_annotation_table.allcelltypes$sample_celltype_name]
pseudobulk_matrix.allcelltypes[1:5,1:4]
```



This code builds a quick TMM normalised matrix. The limma-voom method we used for differential expression worked directly with counts data - but we need something normalised for examining expression! It is customary to log2 transform expression data.

```{r}
dge <- DGEList(pseudobulk_matrix.allcelltypes) 
dge <- calcNormFactors(dge) 
norm_matrix <- edgeR::cpm(dge)                 # Get normalised matrix according to normalisation factors.
lognorm_matrix <- log2(norm_matrix + 1) # add 1 so that zero counts can be logged to 0. 
lognorm_matrix[1:4,1:5]
```



Plotting the most differentially expressed gene, TMSB10, the change is broader than just CD14+ Monocytes.

```{r}
# here we match the sample order by passing the pseudobulk annotation_table sample celltype into the matrix
my_gene <- 'TMSB10'
pseudobulk_annotation_table.allcelltypes$expression <- lognorm_matrix[my_gene, pseudobulk_annotation_table.allcelltypes$sample_celltype_name]


ggplot(pseudobulk_annotation_table.allcelltypes, aes(x=stim, y=expression, fill=stim)) +
  geom_boxplot(outlier.shape=NA) +
  geom_point() +
  theme_bw() +
  facet_wrap(~cell) +
  ggtitle(my_gene)

```



#### Discussion {- .challenge}

Look at the expression of some other genes (differentially expressed or not). Are they globally changing or just in that celltype?  Share an interesting example.

#### {-}



#### What about FindMarkers? {- .rational}

If you look at the seurat [differnetial expression vignette](https://satijalab.org/seurat/articles/de_vignette), you might have seen code that uses 'FindMarkers()' method to look for differential expression between groups.

By default, that function uses a 'wilcox' test to look for differences in gene 
expression, treating _each individual cell as a replicate_.  In cases where biological replicates are simply not available this can be useful (e.g. pilot study, old public dataset or rare sample). 

However, when biological replicates are available, they should be used! 


